Gusto: a toolkit for compatible finite element dynamical cores

### 2023-07-13 source/paper.tex
some progress

\def\MM#1{\boldsymbol{#1}}

@@

\section{Background}

\subsection{Motivation}
\begin{itemize}
\item route to discretisation of eqns with correct properties on non-uniform grids
\item these properties are considered essential for dynamical cores
\item non-uniform grids are considered important for parallel scaling
\item many choices and research to be done therefore flexible implementation is useful
\end{itemize}

Compatible finite elements provide a route to stable, consistent
discretisations of the partial differential equations governing
geophysical fluids which also have the correct wave propagation
properties on non-uniform grids and avoid spurious computational
modes. These properties, along with conservation of mass and
potentially other quantities such as energy and enstrophy, are
considered essential for the accuracy of the dynamical core of weather
and climate models. While the wave propagation and required
conservation properties guide the choice of finite element function
spaces for the discretisation, there are many options for both the
spatial and temporal discretisation and it is not clear which will be
best for any particular problem. Gusto provides a flexible,
easy-to-use toolkit of options for rapid prototyping of different
spatial and temporal discretisations for geophysical fluid dynamics
equations relevant to numerical weather and climate prediction.

In this article we will describe the current capabilities of Gusto and
the design features that enable our flexible implementation whereby
different spatial and temporal discretisations can be applied to a
range of different equations sets relevant for geophysical fluid
modelling. In section \ref{sec: governing} we will outline three
different equation sets, along with their finite element
discretisations. Then in section \ref{sec: design} we will present the
code structure of Gusto followed by a detailed example (section
\ref{sec: FML}) of the Form Manipulation Language that facilitates
automatic manipulation of the finite element forms within our
timestepping code. Section \ref{sec: results} contains a selection of
numerical results that showcases the range of discretisations
available within Gusto. We summarise our current status and future
directions in section \ref{sec: summary}.

\section{Governing Equations}
\label{sec: governing}

The aim of this section is to introduce three governing equation sets
commonly used in the development of weather and climate models, along
with enough detail about the finite element discretiations provided by
Gusto to demonstrate the similarities between them. These similarities
guide the design of the code structure described in the next section
and motivate the use of Form Manipulation Labelling (FML) that
underlies the flexibility of Gusto, a detailed example of which will
be given in section \ref{sec: FML}.

@@

\begin{align}
  \frac{\partial\MM{u}}{\partial t} + \MM{u}\cdot\nabla\MM{u} + f\MM{k}\times\MM{u} + g\nabla (D+b) &= 0, \\
  \frac{\partial D}{\partial t} + \nabla\cdot(\MM{u}D) &= 0,
\end{align}
where $f$ is the Coriolis parameter, $\MM{k}$ is the vertical vector
normal to the domain and $g$ is the acceleration due to gravity. The
nonlinear velocity advection term can be replaced by the sum of a
vorticity-based term and the gradient of the kinetic energy. This is
known as the vector invariant form and it enables the construction of
schemes that conservation energy and/or enstrophy
\citep{mcrae2014energy, bauer2018energy, wimmer2020energy,
  wimmer2021energy}. However, this form can exhibit numerical
instabilities \citep{bell2017numerical} and it is not yet clear which
form is preferable so in Gusto we offer both forms of the equations.

@@

\begin{align}
  \int_\Omega\MM{w}\cdot\frac{\partial\MM{u}}{\partial t}dV + \int_\Omega\MM{w}\cdot(\MM{u}\cdot\nabla)\MM{u} dV + \int_\Omega\MM{w}\cdot f\MM{k}\times\MM{u}dV - \int_\Omega \nabla\cdot\MM{w} g(D+b) dV &= 0, \forall \MM{w}\in\mathbb{V}_1 \\
  \int_\Omega\phi\frac{\partial D}{\partial t}dV + \int_\Omega\phi\nabla\cdot(\MM{u}D) dV &= 0, \forall \phi\in\mathbb{V}_2
\end{align}

\subsection{Compressible Euler Equations}
The dynamical core at the heart of any weather or climate model solves
the rotating compressible Euler equations that model atmospheric flow,
often with additional approximations such as shallow vs deep
atmosphere or hydrostatic vs non-hydrostatic flow. These equations can
be written in terms of

\subsection{Compressible Boussinesq Equations}

\section{Software Design}
\label{sec: design}

@@

\section{Results}
\label{sec: results}

The aim of this section is to present a selection of results that
demonstrate the tools available in Gusto, such as the choice of grids,
different discretisation options, timestepping schemes and
physics. The numerical properties of these configurations
(convergence, stability, accuracy and efficiency) have either been
demonstrated elsewhere, or will be investigated more thoroughly in
future publications.

\begin{itemize}
\item grids
\item FE order
\item spatial discretisation options
\item timestepping schemes
\item physics
\end{itemize}

\subsection{Shallow water simulations}
The standard suite of shallow water test cases originally proposed in
\citet{williamson1992standard}

>>>>

\def\MM#1{\boldsymbol{#1}}
\newcommand{\pp}[2]{\frac{\partial #1}{\partial #2}}

@@

\section{Introduction}

\subsection{Motivation}
\begin{itemize}
\item many choices and research to be done - Colin says so in his Acta Numerica paper! - therefore flexible implementation is useful
\end{itemize}

Compatible finite elements provide a route to stable, consistent
discretisations of the partial differential equations governing
geophysical fluids which also have the correct wave propagation speeds
and avoid spurious computational modes, even on non-uniform
grids. These properties, along with conservation of mass and
potentially other quantities such as energy and enstrophy, are
considered essential for the accuracy of the dynamical core of weather
and climate models \citep{staniforth2012horizontal}. Achieving these
properties on a non-uniform grid is important because of parallel
scaling. While the required properties of the discretisation guide the
choice of finite element spaces, there are many options for both the
spatial and temporal discretisation and it is not clear which will be
best for any particular problem. Gusto provides a flexible,
extensible, easy-to-use toolkit of options for rapid prototyping of
different spatial and temporal discretisations based on the compatible
finite element framework applied to geophysical fluid dynamics
equations relevant to numerical weather and climate prediction. In
this paper we introduce Gusto to the community, summarise its current
capabilities and outline our vision for future development.

\subsection{Background}

\begin{itemize}
\item overall theory is described in \citet{gibson2019compatible, cotter2023compatible}
\item Gusto has grown out of code described in \citet{natale2016compatible, cotter2012mixed, bendall2019recovered, bendall2020compatible, yamazaki2017vertical, shipton2018higher}
\item links to LFRic - we mention the Gung Ho project on the Gusto website so should describe that link here
\end{itemize}

In this article we will describe the current capabilities of Gusto and
the design features that enable our flexible implementation whereby
different spatial and temporal discretisations can be applied to a
range of different equation sets relevant for weather and climate
modelling. In section \ref{sec: governing} we will outline three
different equation sets, along with their finite element
discretisations. Then in section \ref{sec: design} we will present the
code structure of Gusto followed by a detailed example (section
\ref{sec: FML}) of the Form Manipulation Language (FML) that facilitates
symbolic manipulation of the finite element forms within our
timestepping code. Section \ref{sec: results} contains a selection of
numerical results that showcases the range of discretisations
available within Gusto. We summarise our current status and future
directions in section \ref{sec: summary}.

\section{Governing Equations}
\label{sec: governing}
The aim of this section is to introduce three governing equation sets
commonly used in the development of weather and climate models, along
with enough detail about the finite element discretiations provided by
Gusto to demonstrate the similarities between them. These similarities
guide the design of the code structure described in the next section
and motivate the use of Form Manipulation Labelling (FML) that
underlies the flexibility of Gusto, a detailed example of which will
be given in section \ref{sec: FML}. The convention in Gusto is to
write the equations in residual form, meaning that all the terms
appear on the left hand side. We will follow this convention below in
contrast to many papers which present the equations with the pressure
gradient term and any forcing terms on the right hand side.

@@

\begin{align}
  \pp{\MM{u}}{t} + \MM{u}\cdot\nabla\MM{u} + f\hat{\MM{k}}\times\MM{u} + g\nabla (D+b) &= 0, \\
  \pp{D}{t} + \nabla\cdot(D\MM{u}) &= 0,
\end{align}
where $f$ is the Coriolis parameter, $\hat{\MM{k}}$ is the unit vector
pointing upwards and $g$ is the acceleration due to gravity. The
nonlinear velocity advection term can be replaced by the sum of a
vorticity-based term and the gradient of the kinetic energy. This is
known as the vector invariant form and it enables the construction of
schemes that conservation energy and/or enstrophy
\citep{mcrae2014energy, bauer2018energy, wimmer2020energy,
  wimmer2021energy}. However, this form can exhibit numerical
instabilities \citep{bell2017numerical} and it is not yet clear which
form is preferable so in Gusto we offer both forms of the equations.

@@

\begin{align}
  \int_\Omega\MM{w}\cdot\pp{\MM{u}}{t}dV + \int_\Omega\MM{w}\cdot(\MM{u}\cdot\nabla)\MM{u} dV + \int_\Omega\MM{w}\cdot f\hat{\MM{k}}\times\MM{u}dV - \int_\Omega \nabla\cdot\MM{w} g(D+b) dV &= 0, \forall \MM{w}\in\mathbb{V}_1 \\
  \int_\Omega\phi\pp{D}{t}dV + \int_\Omega\phi\nabla\cdot(D\MM{u}) dV &= 0, \forall \phi\in\mathbb{V}_2
\end{align}

\subsection{Compressible Boussinesq Equations}
The compressible Boussinesq equations are commonly used in ocean
modelling. They are useful from a numerical discretisation perspective
because they incorporate compressibility effects without the
complexity of discretising the nonlinear pressure gradient terms that
we shall see in the compressible Euler equations. We write these
equations in terms of the prognostic variables $\MM{u}$, $b$ and
$p$ where $\MM{u}=(u, v, w)$ is now a three dimensional velocity
field, $b$ is the fluid buoyancy and $p$ is pressure. In terms of
these variables, the equations are:

\begin{align}
  \pp{\MM{u}}{t} + 
  (\MM{u}\cdot\nabla)\MM{u} +
  f\hat{\MM{k}}\times \MM{u} + \nabla p + b\hat{\MM{k}} & = 0, \\
  \pp{p}{t} + c^2\nabla\cdot\MM{u} & = 0, \\
  \pp{b}{t} + \MM{u}\cdot\nabla(b) & = 0, \\
\end{align}

\subsection{Compressible Euler Equations}
The dynamical core at the heart of any weather or climate model solves
the rotating compressible Euler equations that model atmospheric flow,
often with additional approximations such as shallow vs deep
atmosphere or hydrostatic vs non-hydrostatic flow (** mention this
here or at the end of the section??). We write these equations in
terms of the prognostic variables $\MM{u}$, $\rho$ and $\theta$ where
$\MM{u}=(u, v, w)$ is now a three dimensional velocity field, $\rho$
is the fluid density and $\theta$ is the potential temperature. In
terms of these variables, the equations are:

\begin{align}
  \pp{\MM{u}}{t} + 
  (\MM{u}\cdot\nabla)\MM{u} +
  f\hat{\MM{k}}\times \MM{u} + c_p\theta\nabla \Pi + g\hat{\MM{k}} & = 0, \\
  \pp{\rho}{t} + \nabla\cdot(\MM{u}\rho) & = 0, \\
  \pp{\theta}{t} + \MM{u}\cdot\nabla\theta & = 0, \\
  \Pi^{(1-\kappa)/\kappa} = \frac{R}{p_0}\rho\theta, & 
\end{align}

\section{Software Design}
\label{sec: design}
Gusto is built on top of the Firedrake package, an automated system
for solving partial differential equations using the finite element
method. This gives many advantages, not least that it provides the
finite element machinery required to translate the discretisations
written in the previous section in to the matrix-vector systems that
we need to solve. This is done using the Unified Form Language (UFL),
a domain specific language that enables the definition of the finite
element forms within Python. The other packages that Firedrake depends
on are summarised in \citet{davies2022towards} ** mention extruded
meshes and PETSc?

@@

\section{Results}
\label{sec: results}
The aim of this section is to present a selection of results that
demonstrate the tools available in Gusto, such as the choice of grids,
order of finite element fucntion spaces, different spatial
discretisation options, timestepping schemes and physics. The
numerical properties of these configurations (convergence, stability,
accuracy and efficiency) have either been demonstrated elsewhere, or
will be investigated more thoroughly in future publications.

\subsection{Shallow water simulations}
The standard suite of shallow water test cases originally proposed in
\citet{williamson1992standard} provides a common starting point for
testing new discretisations. This test suite is usually augmented by
the unstable barotropic jet described in \citet{galewsky2004initial}
(**and linear tests?)
